---
title: "Assignment 5: Data Visualization"
author: "Sam Campbell"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse);library(lubridate);library(here);library(cowplot)
getwd()

NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

NEON_NIWO_Litter_mass_trap_Processed <- read.csv(
  here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE)

#2 
class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)

#change dates from factor to date format
NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate <- ymd(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)

NEON_NIWO_Litter_mass_trap_Processed$collectDate <- ymd(NEON_NIWO_Litter_mass_trap_Processed$collectDate)

#check class again
class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
samtheme <- theme_dark(base_size = 12) +
  theme(axis.text = element_text(color = "blue"),
        legend.position = "left")

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

#first try without separate aesthetics for 2 lakes
#plot4 <- 
  #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                #(aes(x=po4,
                     #y=tp_ug))) +
  #geom_point()
#print(plot4)

#second try with faceting, successful
#plot4 <- 
  #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                #(aes(x=po4,
                     #y=tp_ug))) +
  #geom_point() +
  #facet_wrap(vars(lakename), nrow = 3) +
#print(plot4)

#third try with samtheme
#plot4 <- 
  #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                #(aes(x=po4,
                     #y=tp_ug))) +
  #geom_point() +
  #facet_wrap(vars(lakename), nrow = 3)
  #samtheme
#print(plot4)

#4th try with line of best fit
#plot4 <- 
  #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                #(aes(x=po4,
                     #y=tp_ug))) +
  #geom_point() +
  #geom_smooth(method = lm,
              #color="black") +
  #facet_wrap(vars(lakename), nrow = 3)
  #samtheme
#print(plot4)

#5th version with adjusted axes
#plot4 <- 
  #ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                #(aes(x=po4,
                     #y=tp_ug))) +
  #geom_point() +
  #xlim(0, 50) +
  #geom_smooth(method = lm,
              #color="black") +
  #facet_wrap(vars(lakename), nrow = 3)
  #samtheme
#print(plot4)

#Final version with lake colors instead of facets per John's feedback
plot4 <- 
  ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
                (aes(x=po4,
                     y=tp_ug,
                     color=lakename))) +
  geom_point() +
  xlim(0, 50) +
  geom_smooth(method = lm,
              color="black")
  samtheme
print(plot4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

#Changing class of month from an integer to a factor prior to plot creation
monthnames <- factor(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$month,
       levels = 1:12,
       labels = month.abb)

#boxplot for temperature after change month from integer to factor
plot5a <-
  ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
         aes(x=monthnames,
             y=temperature_C)) +  
  geom_boxplot(aes(color = lakename)) +
  labs(
    title="Temp Plot",
    x="Month") +
  samtheme
print(plot5a)

#boxplot for TP
plot5b <-
  ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
         aes(x=monthnames,
             y=tp_ug)) +  
  geom_boxplot(aes(color = lakename)) +
  labs(
    title="TP Plot",
    x="Month") +
  samtheme
print(plot5b)

#boxplot for TN
plot5c <-
  ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed,
         aes(x=monthnames,
             y=tn_ug)) +  
  geom_boxplot(aes(color = lakename)) +
  labs(
    title="TN Plot",
    x="Month") +
  samtheme
print(plot5c)

#1st try for cowplot combining temperature, TP, and TN; creates large side-by-side plots w/ 3 individual legends
library(cowplot)
cowplot5 <-
plot_grid(plot5a, plot5b, plot5c, nrow = 1, align = "h", axis = "tb")
print(cowplot5)

#extract one legend
legend <- get_legend(plot5a)

#2nd try for cowplot; shows one legend
cowplot5 <- 
  plot_grid(
  plot5a + theme(legend.position="none"),
  plot5b + theme(legend.position="none"),
  plot5c + theme(legend.position="none"),
  legend,
  align = 'vh',
  hjust = -1,
  nrow = 1
)
print(cowplot5)

#Final cowplot with aligned axes
cowplot5 <- 
  plot_grid(
  plot5a + theme(legend.position="none"),
  plot5b + theme(legend.position="none"),
  plot5c + theme(legend.position="none"),
  legend,
  align = 'vh',
  axis = "btlr",
  hjust = -1,
  nrow = 1
)
print(cowplot5)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The 3 variables seem to be aligned over seasons and between lakes, each reaching a peak during the summer months. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

#Plot needle litter dry mass b date with NLCD class color aesthetic
plot6 <-
  ggplot(subset(NEON_NIWO_Litter_mass_trap_Processed, functionalGroup == "Needles"),
         aes(x = collectDate,
             y = dryMass,
             color = nlcdClass)) +
  geom_line()
print(plot6)

#7

#Same plot as 6, but with NLCD classes as facets rather than colors
plot7 <-
  ggplot(subset(NEON_NIWO_Litter_mass_trap_Processed, functionalGroup == "Needles"),
         aes(x = collectDate,
             y = dryMass)) +
  geom_line() +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(plot7)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective. The lines on the plot get jumbled up and hard to see like spaghetti when all displayed on the same plot, but when separated into 3 facets it becomes easier to see the changes in each nlcdClass over time, and to compare the classes across the 3 facets.