---
title: "Assignment 2: Coding Basics"
author: Sam Campbell
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
# I am generating a sequence of numbers from 1 to 100, increasing by fours
sequence_number_one <- seq(1, 100, by = 4)
#2. 
# I am computing the mean and median of my sequence of numbers
mean(sequence_number_one)
median(sequence_number_one)
#3. 
#I am asking R to determine whether the mean is greater than the median
mean(sequence_number_one) > median(sequence_number_one)

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#Student 1 name and grade
sam <- 37
#Did Student 1 pass?
sam > 50
#Student 2 name and grade
john <- 54
#Did Student 2 pass?
john > 50
#Student 3 name and grade
sarah <- 75
#Did Student 3 pass?
sarah > 50
#Student 4 name and grade
lisa <- 44
#Did Student 4 pass? 
lisa > 50
Student_names_and_grades <- data.frame(sam, john, sarah, lisa)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
A matrix can only contain a single class of data. A data frame can contain multiple classes of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'if' and 'else' is the proper option to use here because it determines the outcome based on two different options, whether the test score is 50 or above, or below.
